Cores from kicks: inferring the gravitational recoil history of SMBHs in elliptical galaxies.

### 2024-05-22 papers/paper-corekick/paper-corekick.tex
Update figures and captions

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Cores from kicks]{Cores from kicks: constraints for observations of gravitational recoil in elliptical galaxies}

@@

% Abstract of the paper
\begin{abstract}
\drafting{Investigate the relationship between core size and kick velocity}
\end{abstract}

@@

\subsection{Mass density profiles in three dimensions}
\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{density_3d}
    \caption{
        Three dimensional stellar mass density, with the line colour corresponding to the kick velocity of the SMBH. 
        All simulations with $\vk \geq 0.5v_\mathrm{esc}=900\,\kmps$ share the same colour, as there is little variation in the profiles for these kick velocities. 
        A clear dichotomy is revealed in the mass profiles, whereby high kick velocities induce a core, and low kick velocities do not, with a sudden transition at $\vk=540\,\kmps$.
        }
    \label{fig:density3d}
\end{figure}
Before considering the projected, 2D stellar mass density profiles of the merger remnants, we investigate the 3D stellar mass distribution.
This alleviates the influence of projection effects on inferring if a core has formed during the BH binary merger process, and gives insight into the dependence of how matter is spatially distributed as a function of the recoiling SMBH kick velocity.
\drafting{Maybe move the following two sentences to Atte's section
To centre the data herein, we use the shrinking sphere method \citep{power2003} to determine the stellar centre of mass, and find the time evolution of the CoM does not systematically vary with kick velocity.
Consequently, we are confident that our results are not a result of malign centring.}
For the $\vk=0\,\kmps$ case, we determine the SMBH influence radius to be $r_\mathrm{infl}\simeq 1.5\,\mathrm{kpc}$, and is insensitive to the time of measurement.
In \autoref{fig:density3d}, we observe a clear dichotomy in the 3D spatial distribution of stellar mass, with mergers that had an SMBH with a kick velocity $\vk<540\,\kmps$ having more mass in the central $3\,\mathrm{kpc}$ ($2r_\mathrm{infl}$) than those mergers with $\vk>540\,\kmps$. 
All merger remnants have a consistent mass density distribution at $r>3\,\mathrm{kpc}$. 
Additionally, for kick velocities $\vk<540\,\kmps$, we observe a gradient in decreasing central stellar density with increasing kick velocity.
We check the time dependence of this dichotomy by constructing the 3D stellar mass distributions at additional times $t=0.056\,\mathrm{Gyr}$, $t=0.094\,\mathrm{Gyr}$, and $t=0.12\,\mathrm{Gyr}$, and find the dichotomy persists, with the same kick velocity split.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{density}
    \caption{
        Surface density profiles with 25\% Bayesian HDI for select representative runs.
        Increasing the kick velocity induces a shallower density profile in the inner regions.
        }
    \label{fig:density}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{rb-kick}
    \caption{
        Bayesian estimate of the merger remnant core size $r_b$, scaled to the core size of the pre-merger remnant $r_{\mathrm{b},0}$, as a function of kick velocity $\vk$.
        The core size distributions are shown as box plots, with the median core size indicated by the central mark.
        Larger kick velocities are correlated with larger core sizes. Additionally, a greater spread in the distribution of core sizes over different viewing projections of the merger remnant is associated with larger kick velocities.}
    \label{fig:vkrb}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{rb_pdf}
    \caption{Probability density function of transformation sampled core radius. The SMBH kick velocity is Monte Carlo sampled from the \citet{zlochower2015} relation assuming randomly-aligned azimuthal spins, and pushed through the fitted model in \autoref{eq:vkrb}. The predicted kick velocities range from $\vk=0\,\kmps$ to $\vk\sim4000\,\kmps$.
    The asymmetry at $\rb/r_{\mathrm{b},0}\simeq 1$ is due to the low kick velocities producing a modest core, as seen in \autoref{fig:vkrb}.}
    \label{fig:rb_pdf}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{h4}
    \caption{
        Cumulative $h_4$ parameter, as defined in \autoref{eq:h4}, for the parallel (top panel) and orthogonal (bottom panel) projections.
        Similar to \autoref{fig:density3d}, simulations with $\vk\geq 900\,\kmps$ have a consistent colour.
        Lines terminating with a circle point indicate simulations with $\vk/\kmps < 600$, those with square points indicate simulations with $600 \leq \vk/\kmps < 1200$, those with triangle points $1200 \leq \vk/\kmps < 1800$, and those with diamond points $\vk/\kmps > 1800$.
        For the parallel case, no systematic trend between $h_4$ and $\vk$ appears to exist.
        Conversely, for the orthogonal projection, a dichotomy similar to the bifurcation seen in \autoref{fig:density3d} is observed. 
    }
    \label{fig:h4}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{orbits}
    \caption{
        Orbit analysis of all sixteen merger remnants.
        Stellar particles are assigned to one of seven different orbital families, and binned into ten logarithmically-spaced radial shells such that $0.2 \leq R/\mathrm{kpc} \leq 30.0$.
        Note the colour scheme is consistent with that in \autoref{fig:density3d}, with kick velocities exceeding $900\,\kmps$ represented using the same colour.
    }
    \label{fig:orbits}
\end{figure*}

@@

To explain the presence of the bifurcation in the IFU maps, we must investigate how the stellar density is behaving in the wake of the recoiling SMBH.
We take two representative recoil velocities, $\vk=0\,\kmps$ and $\vk=1500\,\kmps$, and limit our investigation to stellar particles in a rectangular box defined by $0\leq x/\mathrm{kpc}$, $-5\leq y/\mathrm{kpc}\leq 5$, and $-5\leq z/\mathrm{kpc}\leq 5$ (the aperture of the mock IFU maps in \autoref{sec:app_ifu} extended to approximately $5\,\mathrm{kpc}$), recalling that the SMBH is ejected along the positive $x$ axis.
We create velocity vector fields of the stellar particle motion in the $x$-$y$ plane (see \autoref{fig:wake}).
For the $\vk=0\,\kmps$ case, there is an excess of stellar motion parallel to the $x$-axis, with some orthogonal motion present at small radii. 
This is contrasted to the $\vk=1500\,\kmps$ case, where there is no clear alignment of the stellar velocity flow with either the $x$- or $y$- axes.
We interpret this feature as stellar particles diffusing into the potential wake carved by the recoiling SMBH in the $\vk=1500\,\kmps$ case, where the diffusion is neither parallel nor orthogonal to the SMBH motion, but rather at a variety of angles to it.

The IFU maps describe the line of sight velocity distribution, thus the motion along the $x$-axis in \autoref{fig:wake} is equivalent to the parallel projection, and the $y$-axis equivalent to the orthogonal projection previously described in \autoref{ssec:ifu}.
When the SMBH recoil velocity is small, there is no potential wake created by the SMBH, and thus stellar diffusion to fill that wake does not occur, or occurs weakly in the case of a small but non-zero kick velocity.
As a result, stellar velocities are in general aligned with one of the two observation axes, leading to a narrow distribution of stellar velocities.
A result of this is the $h_4$ kinematic moment, which describes excess distribution tails, is close to zero: there are few high velocity outliers to the distribution.
Conversely, when a large SMBH recoil velocity has propelled the SMBH through the galaxy, and stellar particles are diffusing into the potential wake, the variety of incoming angles, coupled with projection effects, results in a wide velocity distribution. 
A result of this is more stellar particles with high velocities, causing high velocity tails to be present in the LOSVD, and hence an increase in the $h_4$ kinematic moment.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{corner_latent_600}
    \caption{
        Corner plot of the latent parameters in the core-S\'ersic model for the $\vk=600\,\kmps$ instance.
        Contours indicate, from dark to light blue, the 25\%, 50\%, 75\%, and 99\% highest density intervals (HDIs), with sample draws beyond the 99\% HDI shown as individual points. 
        All distributions are unimodal with little cross-correlation between the latent variables.
    }
\end{figure*}

\section{Specific IFU maps}\label{sec:app_ifu}
\drafting{Here we include both paralle and orthogonal projections of the IFU maps for three representative kick velocities: $\vk=0\,\kmps$, $\vk=600\,\kmps$, and $\vk=1800\,\kmps$.}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{IFU_para_v0600}
    \caption{
        Mock integral field unit kinematic maps of the $\vk=600\,\kmps$ remnant in the parallel projection, plotted out to $0.25\,\Reff$ in a circular aperture. 
        Note the colour scaling is consistent with \autoref{fig:IFU0p}.
        Most striking is the concentration of negative ordered velocity $V$ slightly offset from the remnant centre.
        The other kinematic moments display a similar spatial offset.
    }
    \label{fig:IFU600p}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{IFU_ortho_v0600}
    \caption{
        The same as \autoref{fig:IFU600p}, but in the orthogonal projection. 
        Note the colour scaling is consistent with \autoref{fig:IFU0p}.
        Compared to the parallel case, the ordered velocity $V$ and $h_3$ moments appear consistent with the $\vk=0,\kmps$ case, however a spatial offset from the centre persists in $\sigma$ and $h_4$.
    }
    \label{fig:IFU600o}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{IFU_para_v1800}
    \caption{
        Mock integral field unit kinematic maps of the $\vk=1800\,\kmps$ remnant in the parallel projection, plotted out to $0.25\,\Reff$ in a circular aperture. 
        Note the colour scaling is consistent with \autoref{fig:IFU0p}.
        This projection displays clear ordered rotation, however there is no clear peak in velocity dispersion.
        The other kinematic moments are quantitatively similar to the $\vk=0,\kmps$ case. 
    }
    \label{fig:IFU1800p}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{IFU_ortho_v1800}
    \caption{
        The same as \autoref{fig:IFU1800p}, but in the orthogonal projection. 
        Note the colour scaling is consistent with \autoref{fig:IFU0p}.
        Aside from the lack of ordered rotation, there is little difference between the orthogonal and parallel projections at this kick velocity. 
    }
    \label{fig:IFU180o}
\end{figure*}

\section{Stellar velocity fields}
We create vector fields of stellar particles for two representative merger remnants, namely $\vk=0\,\kmps$ and $\vk=1500\,\kmps$.
The velocity fields are created for particles in a rectangular box defined by $0\leq x/\mathrm{kpc}$, $-5\leq y/\mathrm{kpc}\leq 5$, and $-5\leq z/\mathrm{kpc}\leq 5$, as described in \autoref{sec:discussion}, and are shown in \autoref{fig:wake}.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{wake}
    \caption{
        Stellar velocity field for $\vk=0\,\kmps$ (left panel) and $\vk=1500\,\kmps$ (right panel), for an aperture that is approximately consistent with the IFU maps presented in e.g. \autoref{fig:IFU0p}. 
        The right panel demonstrates the stellar particles streaming into the potential wake caused by the recoiling SMBH (the SMBH is beyond the extent of the plot), whereas the same motion is not present in the left panel.
        This causes a difference in the LOSVDs of the two systems, noticeably the $h_4$ kinematic moment.
    }
    \label{fig:wake}
\end{figure}

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Cores from kicks]{Cores from kicks: inferring the gravitational recoil history of SMBHs in elliptical galaxies.}

@@

% Abstract of the paper
\begin{abstract}
\drafting{We answer the questions:
\begin{enumerate}
    \item How does a GW kick influence the stellar mass distribution?
    \item Can we infer if a GW kick has occurred in a ETG?
\end{enumerate}
}
\end{abstract}

@@

\subsection{Mass density profiles in three dimensions}
\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{density_3d}
    \caption{
        Three dimensional stellar mass density, with the line colour corresponding to the kick velocity of the SMBH. 
        All simulations with $\vk \geq 0.5v_\mathrm{esc}=900\,\kmps$ share the same colour, as there is little variation in the profiles for these kick velocities. \drafting{Should probably update this if max is 1020, maybe add special colour for 2000?}.
        At radii $R\lesssim3\,\mathrm{kpc}$, the projected stellar mass density decreases smoothly with increasing kick velocity, with the exception of the $\vk=2000\,\kmps$ simulation, in which the SMBH has not returned to the centre of the merger remnant.
        At radii $R>3\,\mathrm{kpc}$ all the density profiles are consistent with each other, indicating that a changing SMBH recoil velocity only affects the inner mass distribution of the galaxy merger remnant.
        }
    \label{fig:density3d}
\end{figure}
Before considering the projected, 2D stellar mass density profiles of the merger remnants, we investigate the 3D stellar mass distribution.
This alleviates the influence of projection effects on inferring if a core has formed during the BH binary merger process, and gives insight into the dependence of how matter is spatially distributed as a function of the recoiling SMBH kick velocity.
\drafting{Maybe move the following two sentences to Atte's section
To centre the data herein, we use the shrinking sphere method \citep{power2003} to determine the stellar centre of mass, and find the time evolution of the CoM does not systematically vary with kick velocity.
Consequently, we are confident that our results are not a result of malign centring.}
For the $\vk=0\,\kmps$ case, \drafting{check} we determine the SMBH influence radius to be $r_\mathrm{infl}\simeq 1.5\,\mathrm{kpc}$.

In \autoref{fig:density3d}, we observe a clear dichotomy in the 3D spatial distribution of stellar mass, with mergers that had an SMBH with a kick velocity $\vk<540\,\kmps$ having more mass in the central $3\,\mathrm{kpc}$ ($2r_\mathrm{infl}$) than those mergers with $\vk>540\,\kmps$. 
All merger remnants have a consistent mass density distribution at $r>3\,\mathrm{kpc}$. 
Additionally, for kick velocities $\vk<540\,\kmps$, we observe a gradient in decreasing central stellar density with increasing kick velocity.
We check the time dependence of this dichotomy by constructing the 3D stellar mass distributions at additional times $t=0.056\,\mathrm{Gyr}$, $t=0.094\,\mathrm{Gyr}$, and $t=0.12\,\mathrm{Gyr}$, and find the dichotomy persists, with the same kick velocity split.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{density}
    \caption{
        Surface density profiles with 25\% Bayesian HDI for select representative runs.
        Increasing the kick velocity induces a shallower density profile in the inner regions, consistent with the three dimensional density maps in \autoref{fig:density3d}.
        }
    \label{fig:density}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{rb-kick}
    \caption{
        Bayesian estimate of the merger remnant core size $r_b$, scaled to the core size of the pre-merger remnant $r_{\mathrm{b},0}$ (left axis) and in physical units (right axis), as a function of kick velocity $\vk$.
        The core size distributions are shown as box plots, with the median core size indicated by the central mark.
        Larger kick velocities are correlated with larger core sizes. Additionally, a greater spread in the distribution of core sizes over different viewing projections of the merger remnant is associated with larger kick velocities.
        Three models, fit to the median $\vk$--$\rb$ trend, are overplotted.
        For the models, we do not show the parameter uncertainty for visual clarity.
        }
    \label{fig:vkrb}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{rb_pdf}
    \caption{Probability density functions of transformation sampled core radius assuming a given model relation. 
    The SMBH kick velocity is Monte Carlo sampled from the \citet{zlochower2015} relation assuming randomly-aligned azimuthal spins, and pushed through each of the three fitted model in \autoref{eq:vkrb} \drafting{add other models}. 
    The sigmoid model is the preferred model, and has its mode at a larger value of $\rb$ than either the exponential or linear models, at $\rb/r_{\mathrm{b},0}\sim2.8$ ($\rb/\mathrm{kpc}\sim1.6$).
    The predicted kick velocities range from $\vk=0\,\kmps$ to $\vk\sim4000\,\kmps$.
    }
    \label{fig:rb_pdf}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{h4}
    \caption{
        Cumulative $h_4$ parameter, as defined in \autoref{eq:h4}, for the parallel (top panel) and orthogonal (bottom panel) projections.
        Similar to \autoref{fig:density3d}, simulations with $\vk\geq 900\,\kmps$ have a consistent colour.
        Lines terminating with a circle point indicate simulations with $\vk/\kmps < 600$, those with square points indicate simulations with $600 \leq \vk/\kmps < 1200$, and the diamond point indicates the $\vk=2000\,\kmps$ simulation.
        For both the parallel and orthogonal projections, for those simulations with $\vk \leq 1020\,\kmps$ (i.e. have their SMBHs settled), there is a systematic decrease in the cumulative $h_4$ parameter with increasing $\vk$ for $\vk\lesssim600\,\kmps$.
        For $600\lesssim\vk/\kmps\leq1020$, the cumulative $h_4$ parameter is indistinguishable from each other. 
        For the $\vk=2000\,\kmps$ case, the cumulative $h_4$ is both positive and significantly larger than the $600\lesssim\vk/\kmps\leq1020$ simulations, for both parallel and orthogonal projections. 
    }
    \label{fig:h4}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{orbits}
    \caption{
        Orbit analysis of the merger remnants, with a consistent colour scheme to \autoref{fig:density3d}.
        Stellar particles are assigned to one of seven different orbital families, and binned into ten logarithmically-spaced radial shells such that $0.2 \leq R/\mathrm{kpc} \leq 30.0$.
        Noticeably, increasing recoil velocity induces a higher fraction of $\pi$-box orbits at all radii, and a complimentary decrease in rosette orbits at radii $r\lesssim1\,\mathrm{kpc}$.
    }
    \label{fig:orbits}
\end{figure*}

@@

To explain the presence of the bifurcation in the IFU maps, we must investigate how the stellar density is behaving in the wake of the recoiling SMBH.
We take two representative recoil velocities, $\vk=0\,\kmps$ and $\vk=1500\,\kmps$, and limit our investigation to stellar particles in a rectangular box defined by $0\leq x/\mathrm{kpc}$, $-5\leq y/\mathrm{kpc}\leq 5$, and $-5\leq z/\mathrm{kpc}\leq 5$ (the aperture of the mock IFU maps in \autoref{sec:app_ifu} extended to approximately $5\,\mathrm{kpc}$), recalling that the SMBH is ejected along the positive $x$ axis.
We create velocity vector fields of the stellar particle motion in the $x$-$y$ plane.
For the $\vk=0\,\kmps$ case, there is an excess of stellar motion parallel to the $x$-axis, with some orthogonal motion present at small radii. 
This is contrasted to the $\vk=1500\,\kmps$ case, where there is no clear alignment of the stellar velocity flow with either the $x$- or $y$- axes.
We interpret this feature as stellar particles diffusing into the potential wake carved by the recoiling SMBH in the $\vk=1500\,\kmps$ case, where the diffusion is neither parallel nor orthogonal to the SMBH motion, but rather at a variety of angles to it.

The IFU maps describe the line of sight velocity distribution, thus the motion along the $x$-axis in \drafting{removed figure} is equivalent to the parallel projection, and the $y$-axis equivalent to the orthogonal projection previously described in \autoref{ssec:ifu}.
When the SMBH recoil velocity is small, there is no potential wake created by the SMBH, and thus stellar diffusion to fill that wake does not occur, or occurs weakly in the case of a small but non-zero kick velocity.
As a result, stellar velocities are in general aligned with one of the two observation axes, leading to a narrow distribution of stellar velocities.
A result of this is the $h_4$ kinematic moment, which describes excess distribution tails, is close to zero: there are few high velocity outliers to the distribution.
Conversely, when a large SMBH recoil velocity has propelled the SMBH through the galaxy, and stellar particles are diffusing into the potential wake, the variety of incoming angles, coupled with projection effects, results in a wide velocity distribution. 
A result of this is more stellar particles with high velocities, causing high velocity tails to be present in the LOSVD, and hence an increase in the $h_4$ kinematic moment.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{corner_latent_600}
    \caption{
        \drafting{NEED TO UPDATE}
        Corner plot of the latent parameters in the core-S\'ersic model for the $\vk=600\,\kmps$ instance.
        Contours indicate, from dark to light blue, the 25\%, 50\%, 75\%, and 99\% highest density intervals (HDIs), with sample draws beyond the 99\% HDI shown as individual points. 
        All distributions are unimodal with little cross-correlation between the latent variables.
    }
\end{figure*}

\section{Specific IFU maps}\label{sec:app_ifu}
\drafting{Here we include both paralle and orthogonal projections of the IFU maps for three representative kick velocities: $\vk=0\,\kmps$, and $\vk=660\,\kmps$. Add others?}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{IFU_para_v0660}
    \caption{
        Mock integral field unit kinematic maps of the $\vk=600\,\kmps$ remnant in the parallel projection, plotted out to $0.25\,\Reff$ in a circular aperture. 
        Note the colour scaling is consistent with \autoref{fig:IFU0p}.
        Most striking is the slightly enhanced ordered velocity $V$ (some $20\,\kmps$) and the ring of negative $h_4$.
    }
    \label{fig:IFU600p}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{IFU_ortho_v0660}
    \caption{
        The same as \autoref{fig:IFU600p}, but in the orthogonal projection. 
        Note the colour scaling is consistent with \autoref{fig:IFU0p}.
        Compared to the parallel case, the ordered velocity $V$ and $h_3$ moments appear consistent with the $\vk=0,\kmps$ case, however the ring of negative $h_4$ persists.
    }
    \label{fig:IFU600o}
\end{figure*}
